---
title: "Indian Artists at the Met Museum"
author: sarika s gupta
format: pdf
editor: visual
---

## Introduction

When I was 12, my family visited NYC. I was excited about going to the Met because I had read "From the Mixed Up Files of Ms. Basil E. Frankweiler." Unfortunately, it was closed the day we visited so we never went inside. Years later when I moved to NYC, my childhood best friend got me a Met membership. I went once a week, sat in what was the children's library and wrote, taking breaks to walk through the museum. The South Asian art area always drew me in and that's where I spent most of my time wandering.

I learned that the Met has over 470,000 artworks in its collection. I wanted to explore how Indian artists are represented across the entire museum (not just in the Asian Art wing). Which departments feature their work? And from what time period(s)? I used the Met's open API to answer my questions.

## Data Collection and Analysis

I used the `httr2` to access the Met Museum API and tidyverse to wrangle and visualize the data.

```{r}
library(httr2)
library(tidyverse)
library(jsonlite)
```

## Searching for Indian Artists

First, I'll search Across the Met collection for objects associated with Indian artists or culture.

Search for Indian artists/culture

```{r}
req_indian <- request("https://collectionapi.metmuseum.org/public/collection/v1/search?artistOrCulture=true&q=indian")

```

Perform the request (send the request and gets the response)

```{r}
resp_indian <- req_indian |> 
  req_perform()
```

Extract and parse JSON data and convert to a df

```{r}
indian_results <- resp_indian |> 
  resp_body_json()
```

Determine number of objects

```{r}
cat("Objects related to 'Indian':", indian_results$total, "\n")
```

This search found 1303 objects. The API's\`artistOrCulture=true\` parameter searched across artist names and cultural origins in all departments, not just Asian Art.

Get object IDs

```{r}
indian_object_ids <- indian_results$objectIDs
```

Getting Detailed Object Info

Rather than fetching all 1303 works (which takes time), I decided to pull a sample of 100 of these works.

```{r}
set.seed(100)
sample_ids <- sample(indian_object_ids, 100)
```

The data shows object IDs, titles, departments, artist information, and dates.

Function to get object details with error handling

```{r}
get_object_details <- function(object_id) {
  Sys.sleep(0.05)
  tryCatch({
    url <- paste0("https://collectionapi.metmuseum.org/public/collection/v1/objects/", object_id)
    req <- request(url)
    resp <- req |> req_perform()
    resp |> resp_body_json()
  }, error = function(e) {
    NULL  # Return NULL if there's an error
  })
}
```

Get details for all sampled objects

```{r}
object_details <- map(sample_ids, get_object_details)
```

Remove any NULLs (failed requests)

```{r}
objects_df <- map_df(object_details, function(obj) {
  tibble(
    objectID = obj$objectID,
    title = obj$title %||% NA,
    department = obj$department %||% NA,
    artistDisplayName = obj$artistDisplayName %||% NA,
    artistNationality = obj$artistNationality %||% NA,
    objectDate = obj$objectDate %||% NA,
    objectBeginDate = obj$objectBeginDate %||% NA,
    medium = obj$medium %||% NA
  )
})
```

Look at data

```{r}
head(objects_df)
```

## Analyzing Department Distribution

Looking to see which departments feature Indian artists/culture. Counting objects by department

```{r}
dept_counts <- objects_df |> 
  count(department, sort = TRUE)
  
dept_counts
```

I was really surprised that the Arms and Armor department contained the most objects by Indian artists/culture (over 50), while the Asian Art department had only 1 work in this sample!

## Visualizing Department Distribution

```{r}
dept_counts |> 
  filter(!is.na(department)) |> 
  ggplot(aes(x = reorder(department, n), y = n)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n), hjust = -0.3, size = 3.5) +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 70, 10)) +
  labs(
    title = "Indian Objects Across Met Museum Departments",
    x = NULL,
    y = "Number of Objects"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))
```

## Conclusion

The data show something surprising: in my sample of 100 Indian objects from the Met, Arms and Armor had the most (over 50 objects), while Asian Art - where I spent so much time wandering - had only 1 object (!). Musical Instruments also had a significant number of Indian artists, along with the Costume Institute and the European Sculpture and Decorative Arts (which was surprising!)

This API Analysis answered my question: Indian artists and objects ARE represented across many departments at the Met, and they are not just confined to the "South Asian art" section. Indian culture shows up in functional objects, weapons, musical instruments, and textiles in addition to traditional art forms (e.g, paintings, sculpture). Next time I visit the Met, I'll wander beyond the South Asian galleries to see if I can find the work of Indian artists in the other departments

---

# EXTENSION BY HAMIDOU BALLO

## Temporal Analysis: When Were These Objects Created?

Building on Gupta's analysis of department distribution, I wanted to explore the temporal dimension: **when were these Indian objects created?** This helps us understand which historical periods are represented in the Met's collection.

### Data Preparation for Temporal Analysis

First, let's examine and clean the date data. The `objectBeginDate` field contains numeric years that we can use for analysis.

```{r}
# EXTENSION: Examine the date data structure
# Filter out objects with valid dates (non-NA and reasonable year values)
temporal_data <- objects_df |>
  filter(!is.na(objectBeginDate),
         objectBeginDate > 0,       # Remove zeros and negative values
         objectBeginDate <= 2024)   # Ensure dates are not in the future

# Display summary of date range
cat("Date range of objects:\n")
cat("Earliest:", min(temporal_data$objectBeginDate), "\n")
cat("Latest:", max(temporal_data$objectBeginDate), "\n")
cat("Number of objects with valid dates:", nrow(temporal_data), "\n")
```

### Creating Century Categories

To better understand temporal patterns, I'll categorize objects by century using tidyverse functions.

```{r}
# EXTENSION: Create century categories for easier visualization
temporal_data <- temporal_data |>
  mutate(
    century = case_when(
      objectBeginDate >= 1900 ~ "20th-21st Century",
      objectBeginDate >= 1800 ~ "19th Century",
      objectBeginDate >= 1700 ~ "18th Century",
      objectBeginDate >= 1600 ~ "17th Century",
      objectBeginDate >= 1500 ~ "16th Century",
      objectBeginDate >= 1400 ~ "15th Century",
      TRUE ~ "Before 15th Century"
    ),
    # Create ordered factor for proper visualization ordering
    century = factor(century, levels = c(
      "Before 15th Century", "15th Century", "16th Century",
      "17th Century", "18th Century", "19th Century", "20th-21st Century"
    ))
  )

# Count objects by century
century_counts <- temporal_data |>
  count(century, sort = FALSE) |>
  arrange(century)

century_counts
```

### Visualizing Temporal Distribution

```{r}
# EXTENSION: Create visualization of objects by time period
ggplot(century_counts, aes(x = century, y = n, fill = century)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "YlOrRd", direction = 1) +
  labs(
    title = "Temporal Distribution of Indian Objects at the Met",
    subtitle = "Most objects date from the 18th-19th centuries",
    x = "Time Period",
    y = "Number of Objects"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(face = "bold", size = 14)
  )
```

### Temporal Distribution by Department

Let's examine which departments have objects from which time periods.

```{r}
# EXTENSION: Cross-tabulate century and department
dept_by_century <- temporal_data |>
  filter(!is.na(department)) |>
  count(department, century) |>
  arrange(department, century)

# Show top combinations
dept_by_century |>
  head(15)
```

### Visualization: Timeline by Department

```{r, fig.width=10, fig.height=6}
# EXTENSION: Create a heatmap-style visualization
dept_by_century |>
  ggplot(aes(x = century, y = reorder(department, n, sum), fill = n)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = n), color = "white", size = 3, fontface = "bold") +
  scale_fill_gradient(low = "steelblue", high = "darkred",
                      name = "Object Count") +
  labs(
    title = "Indian Objects by Department and Time Period",
    subtitle = "Arms and Armor dominated by 18th-19th century pieces",
    x = "Time Period",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )
```

## Extended Conclusion

**Temporal Findings:**

My extension reveals important temporal patterns in the Met's Indian collection:

1. **18th-19th Century Dominance**: The majority of Indian objects in this sample date from the 18th and 19th centuries, coinciding with the height of British colonial presence in India and increased Western interest in Indian art and artifacts.

2. **Arms and Armor Timeline**: The Arms and Armor department's large collection (identified by Gupta) primarily consists of pieces from the 18th-19th centuries, reflecting the period when Mughal and Indian weaponry and armor were highly prized by European collectors.

3. **Limited Ancient Objects**: Very few objects in this sample date before the 15th century, which may reflect both preservation challenges and the Met's particular collecting focus.

4. **Modern Representation**: There are relatively few objects from the 20th-21st centuries in this sample, suggesting the Met's Indian collection emphasizes historical rather than contemporary works.

**Combined Insights**: Together with Gupta's department analysis, we now have a richer picture: Indian culture at the Met is represented not only across multiple departments but also heavily weighted toward the 18th-19th century colonial period, particularly in functional objects like arms, armor, and musical instruments rather than traditional fine art.

---

**Extension completed by:** Hamidou Ballo
**Original analysis by:** Sarika S Gupta
**Date extended:** December 2024
