---
title: "jsDelivr API Vignette"
author: "Masoud Mahdisoltani (original) & Samuel Dosky (extension)"
date: "`r Sys.Date()`"
format: html
editor: source
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  message: false
---

## Introduction

Using `httr2` we can *get* data from the jsDelivr API. jsDelivr is a free CDN for npm and GitHub packages that serves over 150 billion requests per month. Their data API provides detailed information about packages, versions, files, and download statistics. The documentation for this API is located here: [https://www.jsdelivr.com/docs/data.jsdelivr.com](https://www.jsdelivr.com/docs/data.jsdelivr.com).

The question I want to explore is: **How do download statistics and version distributions of popular JavaScript libraries reflect real-world web development trends and challenges?**

## Accessing the API to Analyze Popular JavaScript Packages

The goal of this code will be to explore several popular JavaScript libraries (jQuery, React, Vue.js, Bootstrap, etc.), analyze their download statistics and usage patterns to understand current trends in web development.

First, let's load the necessary libraries and configure the API:

```{r}
#| echo: true
library(httr2)
library(tidyverse)
library(jsonlite)
library(lubridate)
library(scales)

# No API key required
jsdelivr_base_url <- "https://data.jsdelivr.com/v1"

jsdelivr_req <- request(jsdelivr_base_url)
jsdelivr_req
```

Let's pick some of the more popular packages:

```{r}
popular_packages <- c("jquery", "react", "vue", "bootstrap", "lodash")
```

Exploring basic package information for these libraries:

```{r}
#| echo: true

get_package_info <- \(package_name, base_request = jsdelivr_req) {
  response <- base_request %>%
    req_url_path_append("package", "npm", package_name) %>%
    req_perform()
  
  package_data <- resp_body_json(response)
  
  list(
    name = package_name,
    latest_version = package_data$tags$latest %||% "Unknown",
    total_versions = length(package_data$versions),
    versions = head(package_data$versions, 10)
  )
}

package_info_list <- map(popular_packages, \(pkg) {
  Sys.sleep(0.1) # being respectful 
  get_package_info(pkg, jsdelivr_req)
})

package_info <- map(package_info_list, \(pkg) {
  tibble(
    package = pkg$name,
    latest_version = pkg$latest_version,
    total_versions = pkg$total_versions
  ) 
}) %>%
  list_rbind()

package_info
```

Now let's get download statistics for these packages:

```{r}
#| echo: true

get_package_stats <- \(package_name, base_request = jsdelivr_req) {
  response <- base_request %>%
    req_url_path_append("package", "npm", package_name, "stats") %>%
    req_perform()
  
  stats_data <- resp_body_json(response)
  
  list(
    name = package_name,
    rank = stats_data$rank %||% NA,
    type_rank = stats_data$typeRank %||% NA,
    total_hits = stats_data$total %||% 0,
    versions = stats_data$versions %||% list()
  )
}

stats_list <- map(popular_packages, \(pkg) {
  Sys.sleep(0.1)
  get_package_stats(pkg, jsdelivr_req)
})

package_stats <- map(stats_list, \(pkg) {
  tibble(
    package = pkg$name,
    rank = pkg$rank %||% NA,
    type_rank = pkg$type_rank %||% NA,
    total_hits = pkg$total_hits
  )
}) %>%
  list_rbind() %>%
  arrange(rank)

package_stats
```

Then visualize the popularity of these packages:

```{r}
#| echo: true
#| fig-width: 8
#| fig-height: 6

package_stats %>%
  filter(!is.na(total_hits) & total_hits > 0) %>%
  mutate(
    package = fct_reorder(package, total_hits),
    hits_millions = total_hits / 1e6
  ) %>%
  ggplot(aes(x = package, y = hits_millions)) +
  geom_col(fill = "deepskyblue3", alpha = 0.8) +
  geom_text(aes(label = paste0(round(hits_millions, 1), "M")), 
            hjust = 1, size = 3, fontface = "bold", color = "white") +
  coord_flip() +
  labs(
    title = "Total Downloads (Millions) of Popular JavaScript Libraries",
    subtitle = "Via jsDelivr CDN (All Time)",
    caption = "Data source: jsDelivr API"
  ) +
  ggthemes::theme_calc() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.margin = margin(2,20,3,4, "mm"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  ggbreak::scale_y_break(c(2e3,15e3), expand = expansion(0.05, 0))
```

We can get the version distribution patterns for the packages (selecting the top 3):

```{r}
#| echo: true

get_version_details <- \(package_name, base_request = jsdelivr_req) {
  response <- base_request %>%
    req_url_path_append("package", "npm", package_name, 
                        "stats", "version") %>%
    req_perform()
  
  stats_data <- resp_body_json(response)
  
  if (!is.null(stats_data$versions) && length(stats_data$versions) > 0) {
    version_stats <- map(names(stats_data$versions), \(version_name) {
      version_data <- stats_data$versions[[version_name]]
      tibble(
        package = package_name,
        version = version_name,
        hits = version_data$total %||% 0
      )
    }) %>%
    list_rbind()
    
    return(version_stats)
  } else {
    return(tibble(package = package_name, version = character(0), 
                 hits = numeric(0), bandwidth = numeric(0)))
  }
}


key_packages <- package_stats %>% 
  slice_max(total_hits, n = 3) %>% 
  pull(package)

version_details_list <- map(key_packages, \(pkg) {
  Sys.sleep(0.2)
  get_version_details(pkg, jsdelivr_req)
})


all_version_data <- bind_rows(version_details_list) %>%
  filter(hits > 0) %>%
  slice_max(hits, n = 10, by = package)

all_version_data
```

And visualizing version popularity:

```{r}
#| echo: true
#| fig-width: 8
#| fig-height: 5

all_version_data %>%
  mutate(
    hits_millions = hits / 1e6,
    version_short = str_trunc(version, 10)
  ) %>%
  ggplot(aes(x = tidytext::reorder_within(version_short, hits, package), 
             y = hits_millions)) +
  geom_col(aes(fill = package), alpha = 0.8) +
  tidytext::scale_x_reordered() +
  coord_flip() +
  facet_wrap(~package, scales = "free") +
  labs(
    title = "Most Popular Versions by Package",
    subtitle = "Top 10 versions for each library",
    x = "Version",
    y = "Downloads (Millions)",
    caption = "Data source: jsDelivr API"
  ) +
  ggthemes::theme_calc() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

A lot more can be done; we can find the most popular packages, or find the month that more packages see new releases, and much more.

## Conclusion

Bootstrap's download numbers are staggering, over 16 billion downloads completely dwarf jQuery's 1.5 billion, a difference of more than 10x. What's particularly striking is that jQuery, despite being widely considered legacy technology, still maintains such high download volumes; countless production sites continue to rely on jQuery, and migrating away isn't always practical or prioritized.

The version fragmentation across these libraries tells another important story. Rather than seeing all downloads concentrated on the latest versions, we see substantial usage spread across multiple releases. This explains why library maintainers invest so heavily in backwards compatibility, not just within their own package but also with other major packages, knowing that developers often can't or won't upgrade every dependency immediately when new versions are released.

## Extension by Samuel Dosky: Rank vs Total Hits

This extension explores the relationship between a package’s jsDelivr popularity rank
and its total number of downloads. Since a lower rank value corresponds to a more
popular package, we expect packages with lower ranks to have higher total hit counts.
Using a logarithmic scale on the y-axis allows for meaningful comparison between
packages with vastly different download volumes. The visualization confirms that
top-ranked packages dominate total usage, while lower-ranked packages trail
significantly behind.

```{r}

library(dplyr)
library(ggplot2)
library(scales)

package_stats_clean <- package_stats |>
  filter(!is.na(rank), !is.na(total_hits), total_hits > 0)

ggplot(package_stats_clean, aes(x = rank, y = total_hits)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_y_log10(labels = label_number(scale_cut = cut_si("M"))) +
  scale_x_reverse() +
  labs(
    title = "Rank vs Total jsDelivr Hits",
    subtitle = "Lower rank number = more popular package",
    x = "Rank (lower is better)",
    y = "Total Hits (log scale)",
    caption = "Data source: jsDelivr API"
  ) +
  theme_light()
```

